MetaTracts - A Method for Robust Extraction and Visualization of Carbon Fiber Bundles in Fiber Reinforced Composites

### 2014-09-25 chapters/MetaTracts.tex

In this work, we focus on datasets with larger viewports and lower resolutions where the individual carbon fibers (filaments) are indiscernible or are barely visible. Our domain experts are mainly interested in visualizing the geometric structures in the weaving pattern of fiber bundles in endless carbon fiber reinforced composites instead of high resolution studies of individual fibers.
%of short fiber reinforced composites as presented by Weissenbock2014 et al.~\cite{Weissenbock2014}.
Figure~\ref{fig:data-char} provides an example of our targeted dataset. It also shows the recurring fiber bundle pattern in the final CFRP laminate, called a \textit{unit cell}.
Our work is motivated from the recent progress in two interrelated fields: On one hand, CFRP components have gained wide application in a variety of industries  because of its superior material and physical properties in comparison to conventional materials~\cite{Karpat2012}. On the other hand, recent developments of industrial 3D X-ray computed tomography (XCT) with regard to larger detectors, larger field of views, and better resolutions opened XCT for this new application area of non destructive testing for fiber reinforced components~\cite{Schilling2005}.

@@

\begin{figure}[tb]
\centering
%\includegraphics[width=0.45\textwidth]{imagesMT2014/MT_data_4_2.png}
\includegraphics[width=0.5\textwidth, trim = 0mm 110mm 0mm 0mm, clip,]{images_pvis/figure1}
\caption{Data characteristics: (A) shows a rendering of the data. (B) shows 2D slices along the X- and (C) along the Z-axis. (D) close up of the region marked in green in (C), multiple fiber bundles cross and are indistinguishable by visual inspection alone. }
\label{fig:data-char}
\end{figure}
\section {Data Characteristic and Assumptions}
\label{sec:char_data}

Figure \ref{fig:data-char} shows dataset 1 with woven fiber bundles. Dataset 1 is 450$\times$300$\times$500 voxels in size with isotropic resolution of 2$\mu m$, the data type is unsigned 8 bit integer. Figure~\ref{fig:data-char} clearly shows the weaving structure of the composite unit cell as regional subset of the recurring fiber bundle weaving pattern. The unit cell is used as a basis structure for CFRP manufacturing. The alignment and weaving pattern of the fiber bundles in the cloth influences the strength and stiffness properties of the resulting material. Figure~\ref{fig:data-char}a shows a volume rendering of the dataset. Figure~\ref{fig:data-char}b shows 2D slices along the $X$- and Figure~\ref{fig:data-char}c in $Z$-axis. Figure~\ref{fig:data-char}d shows the magnified version of the green region of interest. 
The resolution of dataset 1 renders the individual carbon fibers in a fiber bundle hard to resolve using 2D slice images only. Figure~\ref{fig:data-char}d contains two bundles going in opposite directions. However, the separation between two fiber bundles is barely visible. The carbon fibers themselves do not differ much from the underlying epoxy matrix in terms of attenuation which poses an additional challenge for characterization. The fiber bundles may differ in terms of the amount of fibers in the bundle. Figure~\ref{fig:data-char}a shows the large variation in cross section sizes among the bundles.
Depending on the weaving pattern the fiber bundles cross each other in different orientations. In general the number of orientations is limited by the weaving pattern. Nevertheless, the weaving pattern may cause the individual bundles to be curved. In consequence individual fibers may be adjacent in Euclidean distance but belong to different bundles. 
We make the following assumptions on our data:
\begin{itemize}
\item The structure embedded in the data contains fiber bundles of indiscernible fibers.
\item Local orientation: each point within a fiber bundle has a local orientation which is parallel to the corresponding area in the fiber bundle.
\item The local orientation may change gradually inside the fiber bundle.
\item The local orientation may be noisy and not reliable.
\item Connectivity: moving along the direction of a non-noisy local orientation in small increments, we will reach another neighborhood with similar local orientation.
\item Fiber bundles going in different directions only interact near the surface of contact.
\end{itemize}
%\begin{figure}[htp]
%  \centering
%    \subfloat[Overview of the dataset]{\label{figur:3}\includegraphics[scale=0.3]{imagesMetaTracts/crop13-a.eps}}
%  \subfloat[X Axis]{\includegraphics[width=40mm, height=40mm]{imagesMetaTracts/xaxis.eps}}\label{figur:1}\\
%  \subfloat[Y Axis]{\label{figur:2}\includegraphics[width=40mm, height=40mm]{imagesMetaTracts/yaxis.eps}}
%  \subfloat[Z Axis]{\label{figur:3}\includegraphics[width=40mm, height=40mm]{imagesMetaTracts/zaxis.eps}}
%  \caption{One of our datasets.Fig \ref{fig:orig_data}(a) provides an overview of the entire volume, the rest show slices of each of the major axis.}\label{fig:orig_data}
%\end{figure}

@@

\subsection {MetaTracts generation}
\label{subsec:MetaTracts-generation}
 In this section we explain the process of MetaTracts generation in $\mathbb{R}^2$. The procedure trivially extends to $\mathbb{R}^3$. In $\mathbb{R}^2$ all the above properties hold except that cylinders are replaced by rectangles. The reliable Hessian step  generates a local orientation in $\mathbb{R}^2$ and a $R_H$[0,1]. In Figure \ref{fig:algo} all regions which have reliable Hessian less than the threshold have been marked in blue. Figure~\ref{fig:algo}c shows a single cylinder of a MetaTracts. Let the seed point associated with the MetaTract be grid point $C_{p}$ (property 2 in Section~\ref{subsec:mtprop}). The local orientation at $C_p$ as computed in the step above is $N_P$ and is given by the dark green arrow. We generate a rectangle of length $L$ and radius $R$, the green dashed rectangle shows the rectangle generated at $C_p$ (properties 3 and 5 in Section~\ref{subsec:mtprop}). 
\begin{figure}[tb]
  \begin{center}
 \includegraphics[ width=0.3\textwidth]{imagesMT2014/image4}
  \end{center}
  \vspace{-1em}
 \caption{MetaTracts generation: (B) MetaTract generation, region with $R_H$ less than threshold. (A) close up of marked region in (B). (C) Parts of a single MetaTract.}
  \label{fig:algo}
    \vspace{-1.5em}
\end{figure}
The set of vertices which are in the green region but not overlapped by blue are possible candidates for the start point of the next cylinder in a particular MetaTract. We call these vertices \textit{``candidate vertices"}. From these we select another grid vertex which will be the start point for the next cylinder.
%\begin{figure}
%\centering
% \includegraphics[width=0.25\textwidth]{imagesMetaTracts/angle-diff.png}
%\caption{Angle between the local orientation of two cylinders that are part of the same MetaTract.}
%\label{fig:angle-diff}
%\end{figure}
Based on such assumptions on our data as local orientation and connectivity (Section~\ref{sec:char_data}), the order of the candidate vertices is based on the following characteristics:

\begin{itemize}
\item \textit{Orientation similarity}: We want the orientation of the start points for the consecutive cylinders to have a similar orientations ($N_p$). 
\item \textit{Large distance}: We want the MetaTracts to traverse the data using as few cylinders as possible. Thus, we want the distance between $C_p$ and the start point for the next cylinder to be as large as possible. We measure distance of a candidate vertex from $C_p$ by projecting the Euclidean distance between them onto $N_p$. For example the Euclidean distance between the blue and the orange vertices projected onto $N_P$ gives distance (Figure~\ref{fig:algo}B). We refer to this perpendicular projection distance as \textit{`projected\_dist'}. 
\end{itemize}
% \begin{equation}\label{eqn:algo_1}
% d=\frac{1}{3}e^{(\frac{-angle^{2}}{\alpha^{2}})}+\frac{2}{3}(1-e^{\frac{-{PerpDist}^{2}}{\beta^{2}}})
% \end{equation}
We define a priority for each of the candidate vertices based on the above factors.
For each cylinder in a MetaTract, we put the \textit{candidate vertices} of it in a priority queue based on Equation \ref{eqn:algo_1}.
\begin{equation}
Priority = \gamma_1 e^{(-angle^2 / \alpha^2)} + \gamma_2e^{(-projected\_dist^2 / \beta^2)}
\label{eqn:algo_1}
\end{equation}
$\gamma_1$, $\gamma_2$ are the  weights ($\mathbb{R}_{\ge 0}$)  which decide how the priority depends on the affine combination of the two factors. For all our cases we use $\gamma_1$ = $1 / 3 $ and $\gamma_2$ = $2 / 3$. In general we suggest $\gamma_1 +\gamma_2 = 1 $ and $\gamma_1 \leq \gamma_2$. At each iteration we pick the top element in the priority queue generate the corresponding cylinder and repeat the steps. Essentially, Equation~\ref{eqn:algo_1} selects a grid point which is the furthest from the current point and going in a similar direction. This gives it the advantage of tackling noise/errors in local orientation better than integral curves by looking at multiple choices for vertices and avoiding intra cell interpolation in an already noisy environment.

In this particular example in Figure~\ref{fig:algo}b we select the orange grid vertex next and repeat the process. We do not select the purple vertex (Figure~\ref{fig:algo}c) because it is not a candidate vertex. If we generate tracts that have erroneous local orientations they will not be able to find further possible candidate vertices and will be of short length. Short tracts are then removed. The MetaTracts generated are shown in Figure~\ref{fig:length_distribution}a. The MetaTracts are colored with the mean orientation direction mapped to the RGB space. Figure\ref{fig:length_distribution}a shows a single slice of YZ plane. That consistent orientation is a key intrinsic feature in the data becomes visually pronounced.
We apply uniform, dense seeding to the XCT volume data to trace and generate the tracts of fiber bundles. Figure~\ref{fig:length_distribution}b shows the MetaTracts of a particular bundle colored according to their individual length.
MetaTracts in a given fiber bundle do not extend the full length of the bundle. MetaTracts may have different lengths and may partially overlap.

@@

\subsection{Distance based clustering}
 \label{subsec:dist_clustering}
Orientation clustering uses only orientation information. To subdivide the oriented clusters into fiber bundles, we need to include further information about the geometric proximity between MetaTracts. We use the directed Hausdorff distance for distance based clustering.
Each MetaTract is represented as a set of points ($C_P$). Formally, the directed Hausdorff distance from point set $P$ to point set $Q$ is defined as 
$H_{dir}(P,Q) = max_{p \in P} min_{q \in Q} d(p,q)$ .
The Hausdorff distance is defined as $H(P,Q) = max(H_{dir}(P,Q),H_{dir}(Q,P))$.
The Hausdorff distance is a metric so $H(P,Q) \le H(P,Q') + H(Q',Q)$ but the directed Hausdorff is not.
Unfortunately, the Hausdorff distance does not work well for our application since a fiber bundle may have many MetaTract which only partially covers the bundle (Figure~\ref{fig:length_distribution}b). If a MetaTract $P$ covers only part of the fiber bundle covered by $Q$, then $H_{dir}(P,Q)$ will be very small while $H_{dir}(Q,P)$ will be large.
Thus, $H(P,Q)$ will be large, even though $P$ and $Q$ are in the same fiber bundle.
Instead of using the Hausdorff distance, $\max(H_{dir}(P,Q),H_{dir}(Q,P)$, we use $\min(H_{dir}(P,Q),H_{dir}(Q,P))$. If $P$ covers only part of the fiber bundle covered by $Q$, then $\min(H_{dir}(P,Q),H_{dir}(Q,P))$ is very small.
Note that if $P$ and $Q$ overlap but do not cover the same parts of the fiber bundle, then $H_{dir}(P,Q)$ and $H_{dir}(Q,P)$ and $\min(H_{dir}(P,Q),H_{dir}(Q,P))$ will be large.

@@

In comparison Figure~\ref{fig:comparison} shows the MetaTracts in Figure~\ref{fig:length_distribution}a being hierarchically clustered directly using proximity alone into 10, 15, and 20 clusters (ground truth is 11). As discussed above, single linkage hierarchical clustering in the presence of overlapping fibers tends to create large clusters and small, low-cardinality outliers. When number of clusters ($h$) is 10 two large multi-bundle incorrect clusters are generated and the rest are outliers. As $h$ increases, some appropriate bundles start to form. Even at $h=20$, the top bundles all clusters together into one large incorrectly identified bundle.
\begin{figure}
	\centering
	\includegraphics[width=0.5\textwidth, trim = 0mm 80mm 0mm 170mm, clip]{images_pvis/figure10}
	\caption{MetaTracts data (left) and its straight-forward hierarchical clustering for various numbers of clusters.}
	\label{fig:comparison}
\end{figure}

@@

Apart from direct visualization of the MetaTracts, we show two additional extensions which were requested by our domain specialists as highly important. The first one is voxelizing the original volume according to the clusters each voxel is associated with.
The second is to extract the corresponding surfaces from the voxel data by binarizing the volume per cluster and extracting the isosurface of the largest connected component from the binary volume. Both methods can possibly be used for further analysis of the data, e.g. in simulations.
To voxelize the space,  we compute a neighborhood around each voxel. We then enumerate the number of voxels of each class (cluster) in this neighborhood. The voxel is then assigned to the class with the maximum number of elements in the neighborhood. Figures ~\ref{fig:crop-16-decomp}a and~\ref{fig:prepreg}d show the result of voxelization. Figure~\ref{fig:crop-16-decomp}c,d shows examples of extracted meshes.

@@

 %glass
We tested our method on multiple datasets with varying characteristics. 
Figure~\ref{fig:orientation_clustering} shows the result of dataset~\ref{fig:data-char} decomposed into 2 orientation clusters with each orientation orientation cluster decomposed into $h=10$ clusters (the ground truth is 6 and 5 respectively). The hierarchical clustering is robust to over segmentation.
Figure~\ref{fig:len_dist_crop16} shows the median, minimum and maximum lengths per cluster. We generate the correct clusters and outlier bundles with very few elements which can be discarded. As noted in section~\ref{subsec:dist_clustering} the tracts do not run the entire length of the dataset, the median length of MetaTracts in cluster 6 (Figure~\ref{fig:length_distribution}) is 200 while the maximum MetaTract length in the bundle is about 500 (measured in units of grid cube length: 2 $\mu m$). Figure~\ref{fig:length_distribution}b shows the length per MetaTract distribution for a particular fiber bundle. 
Figure~\ref{fig:prepreg} shows the results on our dataset 2. This dataset is dense with \textit{flat and thin} bundles. The dimensions are 300$\times$350$\times$300 and the datatype is uint16 (dataset 1 is uint8). The bundles are indistinguishable in the original data set Figure~\ref{fig:prepreg}a. A green dotted line shows one of the fiber bundle boundary. Figure~\ref{fig:prepreg}b shows the fiber bundles for the two orientations, Figure~\ref{fig:prepreg}c shows the combined results, and Figure~\ref{fig:prepreg}d shows the volume after the voxelization.
The fiber bundle extraction using MetaTracts was implemented in C++ using ITK. All clustering was done in R. The preprocessing consists of MetaTract generation the distance computation and the orientation clustering. The hierarchical clustering separately, on the results of the orientation clusters can be done withing an order of a few minutes on Intel Xeon E5-2667 workstation according to user input of $h$ (section~\ref{sec:param_choices}). We store the normalized direction at each fiber point, thus both distance measures can be extracted simultaneously without any extra computations.

>>>>

In this work, we focus on datasets with larger viewports and lower resolutions where the individual carbon fibers (filaments) are indiscernible or are barely visible. Our domain experts are mainly interested in visualizing the geometric structures in the weaving pattern of fiber bundles in endless carbon fiber reinforced composites instead of high resolution studies of individual fibers.
%of short fiber reinforced composites as presented by Weissenbock2014 et al.~\cite{Weissenbock2014}.

Figure~\ref{fig:data-char} provides an example of our targeted dataset. It also shows the recurring fiber bundle pattern in the final CFRP laminate, called a \textit{unit cell}.
Our work is motivated from the recent progress in two interrelated fields: On one hand, CFRP components have gained wide application in a variety of industries  because of its superior material and physical properties in comparison to conventional materials~\cite{Karpat2012}. On the other hand, recent developments of industrial 3D X-ray computed tomography (XCT) with regard to larger detectors, larger field of views, and better resolutions opened XCT for this new application area of non destructive testing for fiber reinforced components~\cite{Schilling2005}.

@@

\begin{figure}[tb]
\centering
%\includegraphics[width=0.45\textwidth]{imagesMT2014/MT_data_4_2.png}
\includegraphics[width=0.5\textwidth, trim = 0mm 110mm 0mm 0mm, clip,]{images_pvis/figure1}
\caption{Data characteristics: (A) shows a rendering of the data. (B) shows 2D slices along the X- and (C) along the Z-axis. (D) close up of the region marked in green in (C), multiple fiber bundles cross and are indistinguishable by visual inspection alone. }
\label{fig:data-char}
\end{figure}
\section {Data Characteristic and Assumptions}
\label{sec:char_data}
Figure \ref{fig:data-char} shows dataset 1 with woven fiber bundles. Dataset 1 is 450$\times$300$\times$500 voxels in size with isotropic resolution of 2$\mu m$, the data type is unsigned 8 bit integer. Figure~\ref{fig:data-char} clearly shows the weaving structure of the composite unit cell as regional subset of the recurring fiber bundle weaving pattern. The unit cell is used as a basis structure for CFRP manufacturing. The alignment and weaving pattern of the fiber bundles in the cloth influences the strength and stiffness properties of the resulting material. Figure~\ref{fig:data-char}a shows a volume rendering of the dataset. Figure~\ref{fig:data-char}b shows 2D slices along the $X$- and Figure~\ref{fig:data-char}c in $Z$-axis. Figure~\ref{fig:data-char}d shows the magnified version of the green region of interest. 
The resolution of dataset 1 renders the individual carbon fibers in a fiber bundle hard to resolve using 2D slice images only. Figure~\ref{fig:data-char}d contains two bundles going in opposite directions. However, the separation between two fiber bundles is barely visible. The carbon fibers themselves do not differ much from the underlying epoxy matrix in terms of attenuation which poses an additional challenge for characterization. The fiber bundles may differ in terms of the amount of fibers in the bundle. Figure~\ref{fig:data-char}a shows the large variation in cross section sizes among the bundles.
Depending on the weaving pattern the fiber bundles cross each other in different orientations. In general the number of orientations is limited by the weaving pattern. Nevertheless, the weaving pattern may cause the individual bundles to be curved. In consequence individual fibers may be adjacent in Euclidean distance but belong to different bundles.

We make the following assumptions on our data:
\begin{itemize}[noitemsep]
\item The structure embedded in the data contains fiber bundles of indiscernible fibers.
\item Local orientation: each point within a fiber bundle has a local orientation which is parallel to the corresponding area in the fiber bundle.
\item The local orientation may change gradually inside the fiber bundle.
\item The local orientation may be noisy and not reliable.
\item Connectivity: moving along the direction of a non-noisy local orientation in small increments, we will reach another neighborhood with similar local orientation.
\item Fiber bundles going in different directions only interact near the surface of contact.
\end{itemize}
%\begin{figure}[htp]
%  \centering
%    \subfloat[Overview of the dataset]{\label{figur:3}\includegraphics[scale=0.3]{imagesMetaTracts/crop13-a.eps}}
%  \subfloat[X Axis]{\includegraphics[width=40mm, height=40mm]{imagesMetaTracts/xaxis.eps}}\label{figur:1}\\
%  \subfloat[Y Axis]{\label{figur:2}\includegraphics[width=40mm, height=40mm]{imagesMetaTracts/yaxis.eps}}
%  \subfloat[Z Axis]{\label{figur:3}\includegraphics[width=40mm, height=40mm]{imagesMetaTracts/zaxis.eps}}
%  \caption{One of our datasets.Fig \ref{fig:orig_data}(a) provides an overview of the entire volume, the rest show slices of each of the major axis.}\label{fig:orig_data}
%\end{figure}

@@

\subsection {MetaTracts generation}
\label{subsec:MetaTracts-generation}
 In this section we explain the process of MetaTracts generation in $\mathbb{R}^2$. The procedure trivially extends to $\mathbb{R}^3$. In $\mathbb{R}^2$ all the above properties hold except that cylinders are replaced by rectangles. The reliable Hessian step  generates a local orientation in $\mathbb{R}^2$ and a $R_H$[0,1]. In Figure \ref{fig:algo} all regions which have reliable Hessian less than the threshold have been marked in blue. Figure~\ref{fig:algo}c shows a single cylinder of a MetaTracts. Let the seed point associated with the MetaTract be grid point $C_{p}$ (property 2 in Section~\ref{subsec:mtprop}). The local orientation at $C_p$ as computed in the step above is $N_P$ and is given by the dark green arrow. We generate a rectangle of length $L$ and radius $R$, the green dashed rectangle shows the rectangle generated at $C_p$ (properties 3 and 5 in Section~\ref{subsec:mtprop}).

\begin{figure}[tb]
  \begin{center}
 \includegraphics[ width=0.3\textwidth]{imagesMT2014/image4}
  \end{center}
  \vspace{-1em}
 \caption{MetaTracts generation: (B) MetaTract generation, region with $R_H$ less than threshold. (A) close up of marked region in (B). (C) Parts of a single MetaTract.}
  \label{fig:algo}
    \vspace{-1.5em}
\end{figure}
The set of vertices which are in the green region but not overlapped by blue are possible candidates for the start point of the next cylinder in a particular MetaTract. We call these vertices \textit{``candidate vertices"}. From these we select another grid vertex which will be the start point for the next cylinder.
%\begin{figure}
%\centering
% \includegraphics[width=0.25\textwidth]{imagesMetaTracts/angle-diff.png}
%\caption{Angle between the local orientation of two cylinders that are part of the same MetaTract.}
%\label{fig:angle-diff}
%\end{figure}
Based on such assumptions on our data as local orientation and connectivity (Section~\ref{sec:char_data}), the order of the candidate vertices is based on the following characteristics:

\begin{itemize}
\item \textit{Orientation similarity}: We want the orientation of the start points for the consecutive cylinders to have a similar orientations ($N_p$). 
\item \textit{Large distance}: We want the MetaTracts to traverse the data using as few cylinders as possible. Thus, we want the distance between $C_p$ and the start point for the next cylinder to be as large as possible. We measure distance of a candidate vertex from $C_p$ by projecting the Euclidean distance between them onto $N_p$. For example the Euclidean distance between the blue and the orange vertices projected onto $N_P$ gives distance (Figure~\ref{fig:algo}B). We refer to this perpendicular projection distance as \textit{`projected\_dist'}.

\end{itemize}
% \begin{equation}\label{eqn:algo_1}
% d=\frac{1}{3}e^{(\frac{-angle^{2}}{\alpha^{2}})}+\frac{2}{3}(1-e^{\frac{-{PerpDist}^{2}}{\beta^{2}}})
% \end{equation}
We define a priority for each of the candidate vertices based on the above factors.
For each cylinder in a MetaTract, we put the \textit{candidate vertices} of it in a priority queue based on Equation \ref{eqn:algo_1}.
\begin{equation}
Priority = \gamma_1 e^{(-angle^2 / \alpha^2)} + \gamma_2e^{(-projected\_dist^2 / \beta^2)}
\label{eqn:algo_1}
\end{equation}
$\gamma_1$, $\gamma_2$ are the  weights ($\mathbb{R}_{\ge 0}$)  which decide how the priority depends on the affine combination of the two factors. For all our cases we use $\gamma_1$ = $1 / 3 $ and $\gamma_2$ = $2 / 3$. In general we suggest $\gamma_1 +\gamma_2 = 1 $ and $\gamma_1 \leq \gamma_2$. At each iteration we pick the top element in the priority queue generate the corresponding cylinder and repeat the steps. Essentially, Equation~\ref{eqn:algo_1} selects a grid point which is the furthest from the current point and going in a similar direction. This gives it the advantage of tackling noise/errors in local orientation better than integral curves by looking at multiple choices for vertices and avoiding intra cell interpolation in an already noisy environment.

In this particular example in Figure~\ref{fig:algo}b we select the orange grid vertex next and repeat the process. We do not select the purple vertex (Figure~\ref{fig:algo}c) because it is not a candidate vertex. If we generate tracts that have erroneous local orientations they will not be able to find further possible candidate vertices and will be of short length. Short tracts are then removed. The MetaTracts generated are shown in Figure~\ref{fig:length_distribution}a. The MetaTracts are colored with the mean orientation direction mapped to the RGB space. Figure\ref{fig:length_distribution}a shows a single slice of YZ plane. That consistent orientation is a key intrinsic feature in the data becomes visually pronounced.
We apply uniform, dense seeding to the XCT volume data to trace and generate the tracts of fiber bundles. Figure~\ref{fig:length_distribution}b shows the MetaTracts of a particular bundle colored according to their individual length.

MetaTracts in a given fiber bundle do not extend the full length of the bundle. MetaTracts may have different lengths and may partially overlap.

@@

\subsection{Distance based clustering}
 \label{subsec:dist_clustering}
Orientation clustering uses only orientation information. To subdivide the oriented clusters into fiber bundles, we need to include further information about the geometric proximity between MetaTracts. We use the directed Hausdorff distance for distance based clustering.
Each MetaTract is represented as a set of points ($C_P$). Formally, the directed Hausdorff distance from point set $P$ to point set $Q$ is defined as 
$H_{dir}(P,Q) = max_{p \in P} min_{q \in Q} d(p,q)$ .
The Hausdorff distance is defined as $H(P,Q) = max(H_{dir}(P,Q),H_{dir}(Q,P))$.
The Hausdorff distance is a metric so $H(P,Q) \le H(P,Q') + H(Q',Q)$ but the directed Hausdorff is not.

Unfortunately, the Hausdorff distance does not work well for our application since a fiber bundle may have many MetaTract which only partially covers the bundle (Figure~\ref{fig:length_distribution}b). If a MetaTract $P$ covers only part of the fiber bundle covered by $Q$, then $H_{dir}(P,Q)$ will be very small while $H_{dir}(Q,P)$ will be large.

Thus, $H(P,Q)$ will be large, even though $P$ and $Q$ are in the same fiber bundle.
Instead of using the Hausdorff distance, $\max(H_{dir}(P,Q),H_{dir}(Q,P)$, we use $\min(H_{dir}(P,Q),H_{dir}(Q,P))$. If $P$ covers only part of the fiber bundle covered by $Q$, then $\min(H_{dir}(P,Q),H_{dir}(Q,P))$ is very small.
Note that if $P$ and $Q$ overlap but do not cover the same parts of the fiber bundle, then $H_{dir}(P,Q)$ and $H_{dir}(Q,P)$ and $\min(H_{dir}(P,Q),H_{dir}(Q,P))$ will be large.

@@

In comparison Fig.~\ref{fig:comparison} shows the MetaTracts in Fig.~\ref{fig:length_distribution}(A) being hierarchically clustered directly using proximity alone into 10,15 and 20 clusters (ground truth being 11). As discussed above, single linkage hierarchical clustering in the presence of overlapping fibers tends to create large clusters and small, low cardinality outliers. In Fig.~\ref{fig:comparison} when ``h" is 10 two large multi-bundle incorrect cluster is generated and the rest are outliers. As ``h" increases, some appropriate bundles start to form. Even at ``h" equals 20, the top bundles all clusters together into one large incorrectly identified bundle.

On the other hand, Fig.~\ref{fig:comparison_orient} shows the MetaTracts in Fig.~\ref{fig:length_distribution}(A) being clustered using K-means clustering after being embedded  in a m-dimensional (lower) space using proximity alone into 10, 15, 20 clusters (ground truth being 11). In the lower dimensional space the spatial context is lost and fibers which are in reality far away are grouped together.   
%\begin{figure}[tb]
%\centering
%		\includegraphics[width=0.4\textwidth]{images_pvis/figure7}
%		%\includegraphics[width=0.5\textwidth, trim = 0mm 70mm 0mm 20mm, clip]{imagesMT2014/image7b.pdf}
%	%\includegraphics[width=0.45\textwidth]{imagesMT2014/crop-16/bundles}~ 
%	%\includegraphics[height=0.35\textwidth]{imagesMT2014/crop-16/vol_mesh.png} 
%	\caption{(A) Volume rendering of data set Fig.~\ref{fig:data-char}A.(B) shows a single slice of the data set. (C) shows two of the extracted meshes together. (D) shows the meshes separately.}
%	\label{fig:crop-16-decomp}
%\end{figure}

In comparison Figure~\ref{fig:comparison} shows the MetaTracts in Figure~\ref{fig:length_distribution}a being hierarchically clustered directly using proximity alone into 10, 15, and 20 clusters (ground truth is 11). As discussed above, single linkage hierarchical clustering in the presence of overlapping fibers tends to create large clusters and small, low-cardinality outliers. When number of clusters ($h$) is 10 two large multi-bundle incorrect clusters are generated and the rest are outliers. As $h$ increases, some appropriate bundles start to form. Even at $h=20$, the top bundles all clusters together into one large incorrectly identified bundle.
\begin{figure}
	\centering
	\includegraphics[width=0.5\textwidth]{images_pvis/figure10}
	\caption{MetaTracts data (left) and its straight-forward hierarchical clustering for various numbers of clusters.}
	\label{fig:comparison}
\end{figure}

@@

Apart from direct visualization of the MetaTracts, we show two additional extensions which were requested by our domain specialists as highly important. The first one is voxelizing the original volume according to the clusters each voxel is associated with.
The second is to extract the corresponding surfaces from the voxel data by binarizing the volume per cluster and extracting the isosurface of the largest connected component from the binary volume. Both methods can possibly be used for further analysis of the data, e.g. in simulations.

To voxelize the space,  we compute a neighborhood around each voxel. We then enumerate the number of voxels of each class (cluster) in this neighborhood. The voxel is then assigned to the class with the maximum number of elements in the neighborhood. Figures ~\ref{fig:crop-16-decomp}a and~\ref{fig:prepreg}d show the result of voxelization. Figure~\ref{fig:crop-16-decomp}c,d shows examples of extracted meshes.

@@

 %glass
We tested our method on multiple datasets with varying characteristics.

Figure~\ref{fig:orientation_clustering} shows the result of dataset~\ref{fig:data-char} decomposed into 2 orientation clusters with each orientation orientation cluster decomposed into $h=10$ clusters (the ground truth is 6 and 5 respectively). The hierarchical clustering is robust to over segmentation.
Figure~\ref{fig:len_dist_crop16} shows the median, minimum and maximum lengths per cluster. We generate the correct clusters and outlier bundles with very few elements which can be discarded. As noted in section~\ref{subsec:dist_clustering} the tracts do not run the entire length of the dataset, the median length of MetaTracts in cluster 6 (Figure~\ref{fig:length_distribution}) is 200 while the maximum MetaTract length in the bundle is about 500 (measured in units of grid cube length: 2 $\mu m$). Figure~\ref{fig:length_distribution}b shows the length per MetaTract distribution for a particular fiber bundle. 
Figure~\ref{fig:prepreg} shows the results on our dataset 2. This dataset is dense with \textit{flat and thin} bundles. The dimensions are 300$\times$350$\times$300 and the datatype is uint16 (dataset 1 is uint8). The bundles are indistinguishable in the original data set Figure~\ref{fig:prepreg}a. A green dotted line shows one of the fiber bundle boundary. Figure~\ref{fig:prepreg}b shows the fiber bundles for the two orientations, Figure~\ref{fig:prepreg}c shows the combined results, and Figure~\ref{fig:prepreg}d shows the volume after the voxelization.
The fiber bundle extraction using MetaTracts was implemented in C++ using ITK. All clustering was done in R. The preprocessing consists of MetaTract generation the distance computation and the orientation clustering. The hierarchical clustering separately, on the results of the orientation clusters can be done withing an order of a few minutes on Intel Xeon E5-2667 workstation according to user input of $h$ (section~\ref{sec:param_choices}). We store the normalized direction at each fiber point, thus both distance measures can be extracted simultaneously without any extra computations.

\section {User Evaluation}
\label {sec:user_eval}
\begin{figure}[tb]
	\centering
	\includegraphics[width=0.45\textwidth,  trim = 0mm 60mm 0mm 0mm, clip]{images_pvis/usereval}
	%\includegraphics[width=0.5\textwidth,  trim = 10mm 40mm 0mm 15mm, clip]{imagesMT2014/image_length_histogram.pdf}
	\caption{The user evaluation consists of a set of queries on videos of our results, volume rendering and Hessian colormap. The questionnaire tests the effectiveness of MetaTracts in visualizing ``geometric structure" and ``spatial context". The Likert scale goes from 1-5: strongly disagree, disagree, neutral, agree and strongly agree. }
	\label{fig:userstudy}
\end{figure}
Four NDT practitioners who are familiar with X-Ray computed tomography and its visual representations, four material scientists who are experienced in analyzing fiber reinforced polymers and four practitioners of visualization participated in the user evaluation of our work. We therefore setup a questionnaire with questions on solving given tasks as well as questions using Likert scale (LS) for evaluating the performance of the MetaTracts over currently used techniques. LS consists of 5 levels from ``Strongly disagree" to ``Strongly agree". Fig.~\ref{fig:comparison} shows the results. Task 1 of the questionnaire was to ``identify the number of fiber bundles" from a video showing a 360 degree rotation of the volume rendering of dataset 1 (Fig.~\ref{fig:data-char}A). This task tests the perception of geometric structure and how common problems like uniform gray scale values and occlusion of the different fiber bundles affects the user. As expected, Fig.~\ref{fig:comparison} shows that the participants found it difficult to determine the number of bundles from volume rendering alone. Another common approach is to color map the main eigenvector of Hessian matrix to RGB space. We show a video showing a 360 degree rotation of the volume rendering of the color mapped eigenvectors (similar to Fig.~\ref{fig:reliable_hessian}A) and asked the same query. Compared to the volume rendering, the Hessian color map provides more information about the major orientations and we see that the evaluators performed better in this test than in the direct volume rendering test before. The majority of our participants agreed or strongly agreed (LS) that Hessian color map was more effective. Finally, we showed the MetaTracts extracted bundles (for orientation 1 and 2 separately Fig.~\ref{fig:orientation_clustering}D,E). All participants could correctly identify the number of fiber bundles. All participants agreed that MetaTracts was more effective in perceiving geometric structure than direct volume rendering. They especially liked the 2D slices of the MetaTracts (Fig.~\ref{fig:crop-16-decomp}B, compared to Fig.~\ref{fig:data-char}C).
The next set of tasks on the MetaTracts results tests the spatial context perceived from MetaTracts. We achieve this by asking the evaluators to select the largest fiber bundle from a subset of fiber bundles, to select the smallest from a subset and given a fiber bundle and to select the most similar one from a subset. Due to the problems of occlusions and similar gray scale values these tests are difficult on volume rendering of the plain data only, but straight forward using MetaTracts. In the Likert scale, the majority strongly agreed that the tasks are undemanding compared to volume rendering.
Finally, the majority participants agreed that the extracted meshes provides better spatial context than volume rendering.
